---
title: "penguins_analysis.Rmd"
author: "holly bartlett"
date: "2024-10-16"
output: html_document
---

## Session 1: Cleaning the Palmer Penguins Dataset

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## loading required libraries
library(tidyverse)
library(palmerpenguins)
library(here)
library(janitor)

## loading our function files
source(here("functions", "cleaning.R"))

## loading the raw data
penguins_raw <- read.csv(here("data", "penguins_raw.csv"))

## using our piped function from the functions script - designed to clean the palmer penguins dataset
cleaning_penguin_columns <- function(raw_data){
  print("cleaned names, removed comments, removed empty rows and cols, removed delta")
  raw_data %>% 
    clean_names() %>% 
    shorten_species() %>% 
    remove_empty(c("rows", "cols")) %>% 
    select(-comments) %>% 
    select(-starts_with("delta"))
}

## save the clean data and check the output
colnames(penguins_raw)
penguins_clean <- cleaning_penguin_columns(penguins_raw)
colnames(penguins_clean)

## look at the changes (we can see that species names have been shortened)
head(penguins_clean)


```

Using the renv package to install libraries in a reproducible way

```{r}
# create a lockfile (snapshot) with the current state of dependencies in the project library
renv::snapshot()

# the next user can then run this code to install all the correct libraries (from the snapshot that I created)
#renv::restore() 
```

## Session 2: Exploratory Figures

Loading the packages required for this session (and updating renv)

```{r}
# loading packages
library(ggplot2)
library(patchwork)
library(ragg)
library(svglite)

# updating renv
renv::snapshot()

```

We can make a boxplot as a new variable:

```{r}
flipper_boxplot_demo1 <- ggplot(
  data = penguins_clean, 
  aes(x = species,
      y = flipper_length_mm)) + 
  geom_boxplot()
flipper_boxplot_demo1
```

However, we get a warning message: removed 2 rows containing non-finite outside the scale range (this is because we have some NA values in our dataset)

We want to remove NA values, but only from the columns we are interested in for the boxplot - in this example, we are looking at species and flipper length (mm). We can also make this plot better by adding the (colourblind-friendly) Palmer Penguins colour scheme - and a few other settings.

```{r}
# use piping to subset the columns we want (species and flipper_length_mm) AND remove N/A values
penguins_flippers <- penguins_clean %>% 
  select(species, flipper_length_mm) %>% 
  drop_na() # we already specified the data source at the beginning of the pipe
head(penguins_flippers)

# define colour mapping with names for each species (adelie = darkorange, chinstrap = purple, gentoo = cyan4)
species_colours <- c("Adelie" = "darkorange",
                     "Chinstrap" = "purple",
                     "Gentoo" = "cyan4")

# creating a better boxplot (with no missing values)
flipper_boxplot_demo2 <- ggplot(
  data = penguins_flippers, 
  aes(x = species,
      y = flipper_length_mm)) + 
  geom_boxplot(
    aes(color = species),
    width = 0.3,
    show.legend = FALSE) +  # make data for each species a different colour on the graph (following Palmer penguins colour scheme)
  geom_jitter(aes(color = species),
              alpha = 0.3,  # adds some transparency
              show.legend = FALSE,  # add the raw data on top (jitter gives each point random variation around the central point)
              position = position_jitter(
                width = 0.2,
                seed = 0)) + # random seed: for reproducibility, we want randomness in the figure to be the same every time
  scale_colour_manual(values = species_colours) +
  labs(x = "Penguin species",
       y = "Flipper length (mm)") +  # add axis labels
  theme_bw()  # remove grey background

# view the boxplot we have created 
flipper_boxplot_demo2

```

We can add these functions we have used to a new file ("plotting.R") in our "functions" folder - then load this file:

```{r}
source(here("functions", "plotting.R"))
```

We can create many different plots using this same function (I have named this function "plot_boxplot")

```{r}
# 2 by 2 grid subplots
options(repr.plot.width = 8, repr.plot.height = 8)

# create individual subplots with different y-axis columns
plot1 <- plot_boxplot(penguins_clean, species, flipper_length_mm, "Penguin species", "Flipper length (mm)", species_colours)
plot2 <- plot_boxplot(penguins_clean, species, culmen_length_mm, "Penguin species", "Culmen length (mm)", species_colours)
plot3 <- plot_boxplot(penguins_clean, species, culmen_depth_mm, "Penguin species", "Culmen depth (mm)", species_colours)
plot4 <- plot_boxplot(penguins_clean, species, body_mass_g, "Penguin species", "Body mass (g)", species_colours)

# combine plots in a 2x2 grid
combined_plot <- (plot1 | plot2) / (plot3 | plot4)
combined_plot

```

We often want to save the plots we have created into a folder - here I have created one called "figures". The following code shows how to save plots as png files into this folder.

```{r}

# create a boxplot for the flippers data, using the function specified in "plotting.R"
flipper_boxplot <- plot_boxplot(penguins_clean, species, flipper_length_mm, "Penguin species", "Flipper length (mm)", species_colours)

# adding it as a png to the folder (must run all 3 lines together idk why) 
agg_png("figures/penguin_boxplot_poster.png",
        width = 20, height = 20, units = "cm",  # can set the size of the figure (important for making posters)
        res = 300,  # can change the resolution
        scaling = 1,  # can change how zoomed in the graph is
        snap_rect = TRUE, bitsize = 8)
print(flipper_boxplot)
dev.off()

# we may also want to make a powerpoint figure
agg_png("figures/penguin_boxplot_powerpoint.png",
        width = 20, height = 20, units = "cm", res = 300, scaling = 3,  snap_rect = TRUE, bitsize = 8)
print(flipper_boxplot)
dev.off()

```

We may also want to save plots as svg (vector) files - which are much higher resolution when you zoom in\*\*\*

```{r}
# adding our flippers plot to the folder as an svg file - for some reason this is not giving me a figure!!!
flipper_boxplot <- plot_boxplot(penguins_clean, species, flipper_length_mm, "Penguin species", "Flipper length (mm)", species_colours)
inches_conversion = 2.54
svglite("figures/penguin_boxplot_vector.svg",
        width = 20 / inches_conversion, height = 20 / inches_conversion)
print(flipper_boxplot)
dev.off()

```

I have also added these functions to the "plotting.R" script, in case I want to use them later
